Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-04 application.tex
restructing, moving iso to appendix

The corresponding vertical velocity perturbations are shown in 
Fig. \ref{mmsn_eigenvz}. Although at $r=5$AU $\beta$ depends on $z$,
the disturbance still occupies the fluid column. This is probably due
to the vertically global nature of the fundamental mode. The fact that
we obtain the VSI despite $\beta > \beta_\mathrm{crit}$ near the
midplane suggests, but does not prove, that the VSI can operate
provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
are much longer than the case with $\beta < \beta_\mathrm{crit}$ for
all $z$: we find $t_\mathrm{grow}\sim 40$ orbits at $r=50$AU,
but $t_\mathrm{grow}\sim 400$ orbits at $r=5$AU. Thus, introducing an
optically thick midplane reduces the efficiency of the VSI despite
$\beta < \beta_\mathrm{crit}$ away from the midplane.

>>>>

The corresponding vertical velocity perturbations are shown in 
Fig. \ref{mmsn_eigenvz}. Although at $r=5$AU $\beta$ depends on $z$,
the disturbance still occupies the fluid column. This is probably due
to the vertically global nature of the fundamental mode. The fact that
we obtain the VSI despite $\beta > \beta_\mathrm{crit}$ near the
midplane suggests, but does not prove, that the VSI can operate
provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
are much longer than the case with $\beta < \beta_\mathrm{crit}  \forall z$: we find $t_\mathrm{grow}\sim 40$ orbits at $r=50$AU,
but $t_\mathrm{grow}\sim 400$ orbits at $r=5$AU. Thus, introducing an
optically thick midplane reduces the efficiency of the VSI despite
$\beta < \beta_\mathrm{crit}$ away from the midplane.

### 2015-03-04 linear_adia.tex
restructing, moving iso to appendix

\subsection{Perturbations with thermal relaxation in the thin-disk limit}\label{analytic_relax}
We now introduce thermal relaxation ($\beta>0$), but pursue explicit
solutions in the thin-disk approximation for simplicity. The governing
equation, Eq. \ref{vertiso_gov_nondim}, is then 
\begin{align}
  0 = \delta v_z^{\prime\prime} - z A\delta v_z^\prime +
  (B - C\zhat^2)\delta v_z = 0,\label{nearly_iso_explicit}
\end{align}
with
\begin{align}
  &A \equiv 1 + \ii \epsilon q \hat{k},\\
  &B \equiv \hat{\sigma}^2\left(\chi + \hat{k}^2\right) -
  \left(\chi + \ii \epsilon q \hat{k}\right),\\
  &C \equiv \left(1-\chi\right)\left(\hat{k}^2 - \ii
    \epsilon q\hat{k}\right), 
\end{align}
and $\chi =
\left(1-\ii\hat{\sigma}\beta\right)/\left(1-\ii\hat{\sigma}\beta\gamma\right)
$ in terms of dimensionless variables.

We remark that Eq. \ref{nearly_iso_explicit} is in the same 
form as Eq. 41 in \cite{lubow93}, which describe adiabatic axisymmetric waves in
a vertically isothermal disk without vertical shear. The two equations
become identical after  setting $q=0$ (no vertical shear) and
$\chi=1/\gamma$ (adiabatic flow) in our Eq. \ref{nearly_iso_explicit},
and making the low-frequency, Keplerian disk approximation in
\citeauthor{lubow93} after a change of variables.

\subsubsection{The effect of introducing a small but finite
  thermal relaxation time}\label{relax_pert}
It is instructive to first ask  the more analytically tractable
question: how 
do the eigenfrequencies and eigenfunctions change when we change
$t_c=\beta\Omega_k^{-1}$ from zero to a small but finite value? For
sufficiently small $\beta$, for which the perturbations may be
considered nearly-isothermal, we expect the new solutions to only
differ slightly from the explicit isothermal solutions developed in
\S\ref{iso_poly}.

To determine the (assumed) small effect of introducing a short thermal
relaxation, we linearize Eq. \ref{nearly_iso_explicit} about a known 
solution for $\beta\equiv0$  ($\chi\equiv 1$). 
Let 
\begin{align}\label{nearly_iso_pert}
  \beta \to 0 + \delta\beta,\, \delta v_z\to \delta v_z+\delta
  v_{z1},\,\hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}, 
\end{align}
which implies 
\begin{align}
  \chi \to 1 + \delta\chi = 1 + \ii \hat{\sigma}\left(\gamma-1\right)\delta\beta.
\end{align}
Here $\delta v_z$ corresponds to an explicit solution developed in
\S\ref{iso_poly} with  $\hat{\sigma}^2$ given by Eq. \ref{sig2_iso}. Inserting
Eq. \ref{nearly_iso_pert} into Eq. \ref{nearly_iso_explicit}  and keeping
only first order terms, we obtain 
\begin{align}
 0 =& \delta v_{z1}^{\prime\prime} - z\left(1 + \ii\epsilon q
   k\right)\delta v_{z1}^\prime\notag\\
 &+ \left[2\hat{\sigma}\delta\hat{\sigma}\left(1 + \khat^2\right) +
   \left(\hat{\sigma}^2 - 1\right)\delta\chi \right.\notag\\
&\phantom{++} \left. + \zhat^2\left(\khat^2 -
     \ii\epsilon q \khat\right)\delta\chi\right]\delta v_z\notag\\
 &+\left[\hat{\sigma}^2\left(1 + \khat^2\right) - \left(1+\ii\epsilon
     q \khat\right)\right]\delta v_{z1}.\label{nearly_iso4}
\end{align}

We could proceed by considering the entire set of polynomial solutions
for $W$, and hence for $\delta v_z$, as derived in \S\ref{iso_poly}, but it
is simplest to look at the fundamental mode. Thus we consider the `basic state' 
 \begin{align*}
   \delta v_z = 1,\quad \hat{\sigma}^2\left(1+\hat{k}^2\right) = 1 +
   \ii\epsilon q \hat{k},
 \end{align*}
 then we may seek 
 \begin{align}
   \delta v_{z1} = a \zhat^2 + b,
 \end{align}
where $a$, $b$ are constants.

@@

\subsubsection{Dispersion relation with thermal
  relaxation}\label{disp_relax}
To obtain a dispersion relation similar to that for 
isothermal perturbations (Eq. \ref{sig2_iso}), we first write 
\begin{align}
  \delta v_z(\zhat) =
  g(\zhat)\exp\left(\frac{\alpha\zhat^2}{2}\right), \label{adia_ansatz}
\end{align}
where $\alpha$ is a constant to be chosen for convenience. Inserting
Eq. \ref{adia_ansatz} into Eq. \ref{nearly_iso_explicit} gives
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime + \left(B +
    \alpha\right)g
  +\left(\alpha^2 - \alpha A - C\right)\zhat^2 g.
\end{align}
We choose $\alpha$ to make the coefficient of $\zhat^2g$
vanish, and impose the vertical kinetic energy density
$\rho|\delta v_z|^2\propto |g|^2 \exp{\left(\real\alpha -
    1/2\right)\zhat^2}$ to remain finite. Then assuming $g(\zhat)$
is a polynomial, we require  
\begin{align}
  \real\alpha < \frac{1}{2}, 
\end{align}
which amounts to choosing 
\begin{align}
  \alpha = \frac{1}{2}\left(A - \sqrt{A^2 + 4C}\right).  
\end{align} 
%by considering the case $q=0$ and $\beta\to\infty$ as examined in
%\cite{lubow93}.

We now have an equation in the same form as that for isothermal
perturbations (Eq. \ref{iso_ode3}):
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
which requires
\begin{align}
  B(\hat{\sigma}) + \alpha(\hat{\sigma}) =
  M\left[A-2\alpha(\hat{\sigma})\right].\label{adia_disp0} 
\end{align}
% We remark that for $q=0$ and $\beta\to\infty$, Eq. \ref{adia_disp0} is
% the same as Eq. 54 in \cite{lubow93} when the low-frequency, Keplerian
% approximation is made in the latter. 
% Eq. \ref{nearly_iso_explicit} is same form
% as the governing equation for the adiabatic case
% (Eq. \ref{adia_iso4}). However, the coefficients $\widetilde{B}$ and $\widetilde{C}$
% have a complicated dependence on $\hat{\sigma}$. 
% Applying the same solution method as in \S\ref{adia_explicit}, we 
% find the adiabatic dispersion relation, 
This is a $6^\mathrm{th}$ order 
polynomial in $\hat{\sigma}$:  
\begin{align}
  0 = \sum_{l=0}^{6}c_l\hat{\sigma}^l,\label{relax_disp}
\end{align}
where the coefficients $c_l$ are given in Appendix \ref{relax_coeff}.
The dispersion relation $\hat{\sigma}=\hat{\sigma}(\khat)$ generally
requires a numerical solution. However, for the fundamental mode
($M=0$) some analytic progress can be made, which we consider next.

@@

\subsubsection{Upper limit on the thermal relaxation timescale for the
  fundamental VSI mode}\label{iso_vsi_beta_crit}

Here we estimate the maximum thermal relaxation timescale
$\beta$ that allows the fundamental VSI for $\khat\gg1$.   
% Notice in Fig. \ref{relax_bcrit} the upper limit on $\beta$ for the
% fundamental isothermal VSI tends to a constant for large
% $\khat$. Using this clue, we can calculate this critical $\beta$ by
% considering $\khat\to\infty$. We proceed as follows. 
At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$
is real. The dispersion relation for the fundamental mode is then
\begin{align}\label{relax_disp_fund}
  0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
\end{align} 
Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} and considering 
$\khat\gg1$, we find
\begin{align}
  0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
  \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
  &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
  0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
  - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3.\label{relax_cond2}
\end{align}
This is a pair of simultaneous equations for
$\hat{\omega}$ and $\beta$. % A simple solution can be obtained by the
% following considerations.
Since our discussion of stable solutions  
without vertical shear yield a dispersion
relation $\hat{\omega}\propto \khat^{-1}$ for $\khat\gg 1$
(\S\ref{iso_stable}, Eq. \ref{iso_stable_disp}), we are motivated to
seek solutions to 
Eq. \ref{relax_cond1}---\ref{relax_cond2} with  
% we wish to seek a criterion independent of k 
% \begin{align}
%   \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \hat{\omega} \khat\text{ finite}.
%\end{align}
\begin{align}
  \khat \gg1, \quad |\hat{\omega}|\ll 1 ,\quad
  |\hat{\omega}\khat|=O(1). 
\end{align}
Then
\begin{align}
  \hat{\omega}\khat = \frac{(\gamma-1)\beta}{\epsilon q}
\end{align}
from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
that
\begin{align}
  \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
        q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2} 
\end{align}
at marginal stability for perturbations with large $\khat$. 
For a thin disk $\epsilon\ll 1$, we infer 
\begin{align}\label{iso_vsi_cond}
  \beta \lesssim \frac{|\epsilon q|}{\gamma-1} \equiv
  \beta_\mathrm{crit} 
\end{align}
is required for the fundamental VSI to operate. We will test this
dependence numerically in \S\ref{bcrit_num_test}. 
From our discussion for isothermal perturbations
(\S\ref{iso_discuss}), we expect   
$\beta<\beta_\mathrm{crit}$ to be a sufficient condition for the
VSI. That is, when $\beta<\beta_\mathrm{crit}$, perturbations are
effectively isothermal.

@@

>>>>

\subsection{Thin-disk approximation}\label{analytic_relax}
Our final approximation apply to thin disks for 
which $\epsilon\ll 1$. Then we may write $\rho/\rho_0\simeq
\exp{(-\zhat^2/2)}$. The governing equation,
Eq. \ref{vertiso_gov_nondim}, is then   
\begin{align}
  0 = \delta v_z^{\prime\prime} - z A\delta v_z^\prime +
  (B - C\zhat^2)\delta v_z = 0,\label{nearly_iso_explicit}
\end{align}
with
\begin{align}
  &A \equiv 1 + \ii \epsilon q \hat{k},\\
  &B \equiv \hat{\sigma}^2\left(\chi + \hat{k}^2\right) -
  \left(\chi + \ii \epsilon q \hat{k}\right),\\
  &C \equiv \left(1-\chi\right)\left(\hat{k}^2 - \ii
    \epsilon q\hat{k}\right), 
\end{align}
and $\chi =
\left(1-\ii\hat{\sigma}\beta\right)/\left(1-\ii\hat{\sigma}\beta\gamma\right)
$ in terms of dimensionless variables.

We remark that Eq. \ref{nearly_iso_explicit} is in the same 
form as Eq. 41 in \cite{lubow93}, which describe adiabatic axisymmetric waves in
a vertically isothermal disk without vertical shear. The two equations
become identical after  setting $q=0$ (no vertical shear) and
$\chi=1/\gamma$ (adiabatic flow) in our Eq. \ref{nearly_iso_explicit},
and making the low-frequency, nearly-Keplerian approximation in
\citeauthor{lubow93} after a change of variables.

The thin disk approximation allows us to develop explicit solutions,
but some general results may be established without it (see,
e.g. Appendix \ref{global_corr} and \S\ref{iso_pert}).

\subsection{The effect of introducing a small but finite
  thermal relaxation time}\label{relax_pert}
It is instructive to first ask  the more analytically tractable
question: how 
do the eigenfrequencies and eigenfunctions change when we change
$t_c=\beta\Omega_k^{-1}$ from zero to a small but finite value? For
sufficiently small $\beta$, for which the perturbations may be 
considered nearly-isothermal, we expect the solution to only
differ slightly from a case with $\beta\equiv 0$.

To determine the (assumed) small effect of introducing a short thermal
relaxation, we linearize Eq. \ref{nearly_iso_explicit} about a known 
solution for $\beta\equiv0$, which we choose below. Let  
\begin{align}\label{nearly_iso_pert}
  \beta \to 0 + \delta\beta,\, \delta v_z\to \delta v_z+\delta
  v_{z1},\,\hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}, 
\end{align}
which implies 
\begin{align}
  \chi \to 1 + \delta\chi = 1 + \ii \hat{\sigma}\left(\gamma-1\right)\delta\beta.
\end{align}
Inserting Eq. \ref{nearly_iso_pert} into Eq. \ref{nearly_iso_explicit}
and keeping only first order terms, we obtain  
\begin{align}
 0 =& \delta v_{z1}^{\prime\prime} - z\left(1 + \ii\epsilon q
   k\right)\delta v_{z1}^\prime\notag\\
 &+ \left[2\hat{\sigma}\delta\hat{\sigma}\left(1 + \khat^2\right) +
   \left(\hat{\sigma}^2 - 1\right)\delta\chi \right.\notag\\
&\phantom{++} \left. + \zhat^2\left(\khat^2 -
     \ii\epsilon q \khat\right)\delta\chi\right]\delta v_z\notag\\
 &+\left[\hat{\sigma}^2\left(1 + \khat^2\right) - \left(1+\ii\epsilon
     q \khat\right)\right]\delta v_{z1}.\label{nearly_iso4}
\end{align}

In Eq. \ref{nearly_iso_pert} and \ref{nearly_iso4}, $\delta v_z$ is to
a solution to Eq. \ref{nearly_iso_explicit} with 
$\beta=0$. For definiteness, let us consider the simple solution 
\begin{align}
  \beta\equiv 0, \quad \delta v_z = 1,\quad \hat{\sigma}^2 = \frac{1 +
  \ii\epsilon q \hat{k}}{1+\hat{k}^2}, \label{pert_basic} 
\end{align}
which solves Eq. \ref{nearly_iso_explicit} because $\delta
v_z=$constant, $\chi=1$ so that $C=0$, and the eigenfrequency makes
$B=0$. Then we may seek 
\begin{align}
  \delta v_{z1} = a \zhat^2 + b,
\end{align}
where $a$, $b$ are constants.

% We could proceed by considering the entire set of polynomial solutions
% for $W$, and hence for $\delta v_z$, as derived in \S\ref{iso_poly}, but it
% is simplest to look at the fundamental mode.

@@

\subsection{Explicit solutions and dispersion relation}\label{disp_relax}
%To obtain a dispersion relation similar to that for 
%isothermal perturbations (Eq. \ref{sig2_iso}), 
We now solve Eq. \ref{nearly_iso_explicit} explicitly. We first write  
\begin{align}
  \delta v_z(\zhat) =
  g(\zhat)\exp\left(\frac{\alpha\zhat^2}{2}\right), \label{adia_ansatz}
\end{align}
where $\alpha$ is a constant to be chosen for convenience. Inserting
Eq. \ref{adia_ansatz} into Eq. \ref{nearly_iso_explicit} gives
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime + \left(B +
    \alpha\right)g
  +\left(\alpha^2 - \alpha A - C\right)\zhat^2 g.
\end{align}
We choose $\alpha$ to make the coefficient of $\zhat^2g$
vanish, and impose the vertical kinetic energy density
$\rho|\delta v_z|^2\propto |g|^2 \exp{\left(\real\alpha -
    1/2\right)\zhat^2}$ to remain finite as $|\zhat|\to\infty$. 
Then assuming $g(\zhat)$ is a polynomial, we require  
\begin{align}
  \real\alpha < \frac{1}{2}, 
\end{align}
which amounts to choosing 
\begin{align}
  \alpha = \frac{1}{2}\left(A - \sqrt{A^2 + 4C}\right).  
\end{align} 
%by considering the case $q=0$ and $\beta\to\infty$ as examined in
%\cite{lubow93}.

Eq. \ref{nearly_iso_explicit} becomes 
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
which requires
\begin{align}
  B(\hat{\sigma}) + \alpha(\hat{\sigma}) =
  M\left[A-2\alpha(\hat{\sigma})\right].\label{adia_disp0} 
\end{align}
% We remark that for $q=0$ and $\beta\to\infty$, Eq. \ref{adia_disp0} is
% the same as Eq. 54 in \cite{lubow93} when the low-frequency, Keplerian
% approximation is made in the latter. 
% Eq. \ref{nearly_iso_explicit} is same form
% as the governing equation for the adiabatic case
% (Eq. \ref{adia_iso4}). However, the coefficients $\widetilde{B}$ and $\widetilde{C}$
% have a complicated dependence on $\hat{\sigma}$. 
% Applying the same solution method as in \S\ref{adia_explicit}, we 
% find the adiabatic dispersion relation, 
This is a $6^\mathrm{th}$ order 
polynomial in $\hat{\sigma}$:  
\begin{align}
  0 = \sum_{l=0}^{6}c_l\hat{\sigma}^l,\label{relax_disp}
\end{align}
where the coefficients $c_l$ are given in Appendix \ref{relax_coeff}.
The dispersion relation $\hat{\sigma}=\hat{\sigma}(\khat)$ generally
requires a numerical solution. However, for the fundamental mode
($M=0$) some analytic progress can be made, which we consider next.

@@

\subsection{Upper limit on the thermal relaxation timescale for the
  fundamental VSI mode}\label{iso_vsi_beta_crit}

Here we estimate the maximum thermal relaxation timescale
$\beta$ that allows the fundamental VSI for $\khat\gg1$.   
% Notice in Fig. \ref{relax_bcrit} the upper limit on $\beta$ for the
% fundamental isothermal VSI tends to a constant for large
% $\khat$. Using this clue, we can calculate this critical $\beta$ by
% considering $\khat\to\infty$. We proceed as follows. 
At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$
is real. The dispersion relation for the fundamental mode is then
\begin{align}\label{relax_disp_fund}
  0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
\end{align} 
Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} and considering 
$\khat\gg1$, we find
\begin{align}
  0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
  \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
  &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
  0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
  - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3.\label{relax_cond2}
\end{align}
This is a pair of simultaneous equations for
$\hat{\omega}$ and $\beta$.  
% Since our discussion of stable solutions without vertical shear yield
% a dispersion relation $\hat{\omega}\propto \khat^{-1}$ for $\khat\gg
% 1$ (\S\ref{iso_stable}, Eq. \ref{iso_stable_disp}).
We seek solutions to 
Eq. \ref{relax_cond1}---\ref{relax_cond2} with  
% we wish to seek a criterion independent of k 
% \begin{align}
%   \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \hat{\omega} \khat\text{ finite}.
%\end{align}
\begin{align}
  \khat \gg1, \quad |\hat{\omega}|\ll 1 ,\quad
  |\hat{\omega}\khat|=O(1). 
\end{align}
Then
\begin{align}
  \hat{\omega}\khat = \frac{(\gamma-1)\beta}{\epsilon q}
\end{align}
from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
that
\begin{align}
  \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
        q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2} 
\end{align}
at marginal stability for perturbations with large $\khat$. 
For a thin disk $\epsilon\ll 1$, we infer 
\begin{align}\label{iso_vsi_cond}
  \beta \lesssim \frac{|\epsilon q|}{\gamma-1} \equiv
  \beta_\mathrm{crit} 
\end{align}
is required for the fundamental VSI to operate. We will test this
dependence numerically in \S\ref{bcrit_num_test}. 
From our discussion in \S\ref{relax_pert} we expect   
$\beta<\beta_\mathrm{crit}$ is required for the fundamental mode to be
unstable. That is, when $\beta<\beta_\mathrm{crit}$, perturbations are 
effectively isothermal.

@@

\subsection{Isothermal perturbations}\label{iso_pert}
